---
title: "Working with United States Census Data Using tidycensus"
output: html_document
---


```{r, echo=FALSE,results=FALSE,message=FALSE}
library(tidycensus)
library(tidyverse)
library(knitr)
Sys.getenv("CENSUS_API_KEY")
```

```{r, eval=FALSE}
library(tidycensus)
library(tidyverse)
census_api_key("YOUR API KEY GOES HERE")
```

## Inspect variables for Decennial Census
```{r view variables}
decennial_variables<-load_variables(2010,"sf1")
View(decennial_variables)
```

```{r,echo=FALSE}
kable(head(decennial_variables))
```


## Calling and Manipulating Data using tidycensus and dplyr: Basics

### Calling and Viewing Data

Let's find out the population, by state, in 2010: 

```{r, message=FALSE}
state_population_2010<-get_decennial(geography = "state", 
                                                  variables = "P001001", 
                                                  year = 2010)
View(state_population_2010)
```


```{r,echo=FALSE}
kable(head(state_population_2010))
```

We can adjust the geography and year parameters; let's say we want the population distribution 
across CO counties in the year 2010:

```{r,message=FALSE}
CO_county_population_2010<-get_decennial(geography = "county", 
                                         state="CO",
                                         variables = "P001001", 
                                         year = 2010)
View(CO_county_population_2010)
```

```{r,echo=FALSE}
kable(head(CO_county_population_2010))
```

### Cleaning and Manipulating Data Using dplyr

Let's clean up the table by removing the "variable" column, and renaming the "value" column as "Population", which we can do using functions from the "dplyr" package. 

```{r, message=FALSE}
CO_county_population_2010<-get_decennial(geography = "county", 
                                         state="CO",
                                         variables = "P001001", 
                                         year = 2000) %>% 
                           mutate(variable=NULL) %>% 
                           rename(population=value)

View(CO_county_population_2010)
```

```{r,echo=FALSE}
kable(head(CO_county_population_2010))
```

It's also possible to call multiple variables into a single table. To see this, let's add a field/column containing the rural population in each state in 2010 (as well as the total population in that year), and order the dataset in descending order with respect to the rural population (such that the state with the largest rural population will appear as the first record in the dataset):

```{r,message=FALSE,results=FALSE}
state_pop_ruralpop_2010<-get_decennial(geography = "state", 
                                          variables = c("P001001", "P002005"),
                                          output="wide",
                                          year = 2010) %>% 
                                          rename(total_population=P001001, rural_population=P002005) %>% 
                                          arrange(desc(rural_population))

state_pop_ruralpop_2010

```

```{r,echo=FALSE}
kable(head(state_pop_ruralpop_2010))
```

Let's generate a new variable based on the variables we already have in the dataset. This variable will measure the percentage of each state's population that are rural residents (calculated by dividing the rural population by the total population, and multiplying by 100). We'll call this variable "rural_pct". We'll also resort the dataset, so that it's sorted in descending order with respect to the new "rural_pct" variable, instead of the actual number of rural residents:


```{r}
state_pop_ruralpop_2010<-
  state_pop_ruralpop_2010 %>% mutate(rural_pct=(rural_population/total_population)*100) %>% 
                              arrange(desc(rural_pct))
  
View(state_pop_ruralpop_2010)
```

```{r,echo=FALSE}
kable(head(state_pop_ruralpop_2010))
```




##Visualization and mapping using ggplot and tmap

Let's 


## Appendix 

```{r}
my_years<-c(2000,2010)
population_rural_2000_2010<-map(
  my_years,
  ~(get_decennial(geography = "state", 
                 variables = c("P001001", "P002005"),
                 output="wide",
                 year =.)) %>% 
      mutate(rural_pct=(P002005/P001001)*100) %>% 
      arrange(NAME)
  )

names(population_rural_2000_2010)<-my_years

joined_ds<-full_join(population_rural_2000_2010[["2000"]],population_rural_2000_2010[["2010"]],by="NAME") %>% 
           mutate(pct_change=rural_pct.y-rural_pct.x) %>% 
           select(NAME,pct_change)
joined_ds
```

