---
title: "Census Data in R"
author: "Aditya Ranganath"
date: "2/17/2022"
output:
  html_document:
        toc: true
        toc_depth: 3
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preliminaries: Load Libraries

```{r, message=F, results=F, warning=F}
library(tidycensus)
library(tidyverse)
library(sf)
library(tmap)
```

# Census Data Extraction: Basics

## Extracting a table of variable codes

```{r}
# Calls up variable codes from 2010 decennial census and assigns to object named "decennial_variables"
decennial_variables<-load_variables(2010,"sf1")
```

```{r}
# Prints contents of "decennial_variables"
decennial_variables
```

## Extracting data with ```get_decennial``` 

### Example 1

```{r, message=F, warning=F, results=F}
# Generate a table of population by state, based on the 2010 decennial census
state_population_2010<-get_decennial(geography = "state", 
                                     variables = "P001001", 
                                     geometry=TRUE,
                                     year = 2010)
```

```{r}
# prints "state_population_2010"
state_population_2010
```

### Example 2

```{r, results=FALSE, warning=FALSE, message=FALSE}
# Generate a table of the population distribution across Colorado counties in the year 2010, based on the 2010 decennial census

CO_county_population_2010<-get_decennial(geography = "county", 
                                         state="CO",
                                         variables = "P001001", 
                                         year = 2010)
```

```{r}
# Prints "CO_county_population_2010"
CO_county_population_2010
```

## Cleaning Extracted Data

```{r}
# Clean up the table of CO population by county by removing the "variable" column, and renaming the "value" column as "Population", which we can do using the tidyverse's "dplyr" package

CO_county_population_2010<-CO_county_population_2010 %>% 
                           mutate(variable=NULL) %>% 
                           rename(population=value)

```

```{r}
# prints updated contents of "CO_county_population_2010"
CO_county_population_2010
```

## Extracting Multiple Census Variables

```{r}
# Create a new object containing a dataset of rural population AND overall population by state,based on the 2010 decennial. Also, renames the variables, and arrange the dataset in descending order with respect to the rural population

state_pop_ruralpop_2010<-get_decennial(geography = "state", 
                                       variables = c("P001001", "P002005"),
                                       output="wide",
                                       year = 2010) %>% 
                          rename(total_population=P001001, rural_population=P002005) %>% 
                          arrange(desc(rural_population))
```


```{r}
# prints contents of "state_pop_ruralpop_2010"
state_pop_ruralpop_2010
```

## Creating New Variables

```{r}
#Using dplyr, generate a new variable in the dataset based on existing variables; 
# in particular, create a variable called "rural_pct" that reflects the rural population 
# as a percentage of the population, and then arranges it in descrend order
# with respect to the new "rural_pct" variable

state_pop_ruralpop_2010<- state_pop_ruralpop_2010 %>% 
                            mutate(rural_pct=(rural_population/total_population)*100) %>% 
                            arrange(desc(rural_pct))

```

```{r}
# prints "state_pop_ruralpop_2010"
state_pop_ruralpop_2010
```

## Filtering census datasets

```{r}
# Extracts observations from "state_pop_ruralpop_2010" where rural_pct>40 and assigns to a new object named "rural_pct_over40"

rural_pct_over40<-state_pop_ruralpop_2010 %>%
                  filter(rural_pct>40)
```

```{r}
rural_pct_over40
```

**Student Exercise 1: Create a dataset of Colorado counties whose rural population percentage(with respect to the overall county population) exceededs 50% (based on the 2010 decennialcensus). Sort the dataset in descending order with respect to the rural percentage variable.**

# Census Data Visualization

## Using ggplot to visualize census data

Make a graph that visually conveys the median age in Colorado, by county, based on the 2010 census.

```{r, message=F, warning=F, results=F}
# Extracts Colorado median age dataset
median_age_CO<- get_decennial(geography = "county",
                              state="CO",
                              variables = "P013001", 
                              geometry=TRUE,
                              year = 2010) %>% 
                rename(median_age=value) %>% 
                mutate(County=str_remove(NAME, " County, Colorado")) %>% 
                select(-NAME)
```


```{r}
median_age_CO
```

```{r, fig.width=10, fig.height=10}
# Creates ggplot visualization of CO median age
median_age_CO__visualization<-
  median_age_CO %>%
  ggplot(aes(x = median_age, y = reorder(County, median_age))) + 
  geom_point()+
  labs(title="Median Age by County, CO", x="Median Age", y="County", caption="Source: United States census accessed via tidycensus")+
  theme(plot.title=element_text(hjust=0.5),
        plot.caption=element_text(size=5))
```


```{r, fig.width=10, fig.height=10}
median_age_CO__visualization
```

## Using *tmap* to visualize census data

### Static Map

```{r}
# Makes map object of median age
median_age_CO_map<-
  tm_shape(median_age_CO)+
      tm_polygons(col="median_age",
                  breaks=c(30,35,40,45,50),
                  palette="YlGnBu", 
                  midpoint=TRUE)+
       tm_layout(frame=FALSE, 
                 main.title="Median Age by County,\nColorado",  
                main.title.position="left", 
                legend.outside=TRUE,
                attr.outside=TRUE)+
      tm_credits("Source: US Census via tidycensus", position=c("right", "bottom"))
```

```{r}
median_age_CO_map
```

### Webmap 

```{r}
# changes tmap_mode to "View"
tmap_mode("view")
```

```{r}
# prints "median_age_CO_map" in "View" mode
median_age_CO_map
```

```{r}
# changes mode to "lot"
tmap_mode("plot")
```

```{r}
# prints "median_age_CO_map" in "Plot" mode
median_age_CO_map
```

# Extracting American Community Survey data

```{r}
# Extract 5-year ACS data variable codes for year ending in 2018
ACS_5_2018<-load_variables(2018,"acs5")
```

```{r}
# prints "ACS_5_2018"
ACS_5_2018
```

```{r}
# Uses 2018 ACS to extract median income estimates by county for CO
median_income_CO_counties_2018<-get_acs(geography="county",
                                     state="CO",
                                     variables="B19013_001",
                                     year=2018) %>% 
                            rename(median_income=estimate) %>% 
                            arrange(desc(median_income))
```


```{r}
median_income_CO_counties_2018
```


# ACS Data Visualization

```{r}
# Creates point visualization of median income by county with error bars
median_income_CO_counties_2018_viz<-
     median_income_CO_counties_2018 %>% 
        mutate(County_Name=str_remove_all(NAME,"County, Colorado")) %>% 
          ggplot(aes(x=median_income,y=reorder(County_Name, median_income)))+
           geom_errorbarh(aes(xmin = median_income - moe, xmax = median_income + moe)) +
           geom_point(color = "blue", size = 3)+
                      labs(title="Median Income in Colorado, by County (2018)",
                              y="", x="Median Income Estimate from 5 year ACS\n(Bars indicate margin of error)")+
                             theme(plot.title=element_text(hjust=0.5))
```


```{r, fig.width=10, fig.height=10}
# prints "median_income_CO_counties_2018_viz"
median_income_CO_counties_2018_viz
```







